---
title: "Advanced Economies Labour Market Dashboard"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
    logo: assets/bgi-logo.svg
    css: assets/dashboard.css
    self_contained: true
    source_code: embed
    navbar:
      - { title: "About", href: "README.md", align: right }
---

```{r setup, include=FALSE}
# Load packages, source the data pipeline, and pre-compute derived series.
# All visualisations downstream read from the named list produced here —
# no module is re-queried inside individual chart chunks.
library(flexdashboard)
library(tidyverse)
library(lubridate)
library(scales)
library(here)
library(knitr)
library(DT)

# BGI palette hardcoded so the dashboard renders without the private `bgi`
# R package (e.g. on a GitHub Actions runner). Hex codes are taken directly
# from bgi::bgi_color_universe / scale_color_bgi_d("Default") — the first
# seven entries reproduce the BGI "Default" categorical scale; the remaining
# entries are pulled from bgi_color_universe to extend the cycle so charts
# with up to ~14 distinct categories (e.g. NACE-section small multiples)
# render without "Insufficient values in manual scale" errors.
bgi_palette_default <- c(
  "#e0732b", # orange
  "#4A71A0", # teal
  "#FAB619", # yellow
  "#666699", # purple-grey
  "#4CAF50", # green
  "#d15868", # red
  "#42769b", # blue
  "#2D5178", # dark teal
  "#A44914", # dark orange
  "#c22036", # dark red
  "#03497a", # dark blue
  "#58585B", # grey
  "#efb995", # light orange
  "#729ACB"  # light teal
)
bgi_orange_green   <- c(low = "#A44914", mid = "#FEFDFC", high = "#2E7D32")
# Use colorRampPalette as a final safety net so any n is supported.
bgi_palette_fn     <- function(n) {
  if (n <= length(bgi_palette_default)) bgi_palette_default[seq_len(n)]
  else colorRampPalette(bgi_palette_default)(n)
}
scale_color_bgi    <- function(...) discrete_scale("colour", palette = bgi_palette_fn, ...)
scale_fill_bgi     <- function(...) discrete_scale("fill",   palette = bgi_palette_fn, ...)
scale_color_bgi_og <- function(midpoint = NULL, ...) {
  scale_color_gradient2(low = bgi_orange_green["low"],
                        mid = bgi_orange_green["mid"],
                        high = bgi_orange_green["high"],
                        midpoint = midpoint, ...)
}

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 96
)

# Source the master loader. load_all_data() will read the per-module RDS
# caches in data/ when they are <24h old, so this is cheap on repeat builds.
source(here("R", "load_data.R"))
lm_data <- load_all_data()

# Pull out individual tibbles for readability
vacancy_rate <- lm_data$vacancy_rate
unemployment <- lm_data$unemployment
employment   <- lm_data$employment
activity     <- lm_data$activity
emp_nace     <- lm_data$emp_nace
emp_cob      <- lm_data$emp_cob

# Country ordering used everywhere — keeps legends and facets stable
country_order <- c("United States", "Germany", "France", "Italy",
                   "Spain", "Netherlands", "United Kingdom")

# ---------------------------------------------------------------
# Derived series (computed once, reused across panels)
# ---------------------------------------------------------------

# Latest-quarter snapshot per country for the KPI table on the Overview tab
latest_snapshot <- function(df, value_col, age_filter = NULL) {
  if (nrow(df) == 0) return(tibble())
  if (!is.null(age_filter) && "age" %in% names(df)) {
    df <- df %>% filter(age == age_filter)
  }
  df %>%
    group_by(country_name) %>%
    filter(date == max(date, na.rm = TRUE)) %>%
    ungroup() %>%
    select(country_name, date, {{ value_col }})
}

# Age choices reflect coverage of the underlying sources:
#  - Eurostat publishes both Y15-64 and Y20-64 for employment/activity;
#  - US (FRED OECD-harmonised) and UK (ONS) only publish Y15-64.
# Use Y15-64 throughout for full 7-country coverage.
kpi_unemp <- latest_snapshot(unemployment, u_rate,   "Y15-74") %>%
  rename(unemployment_pct = u_rate, unemp_date = date)
kpi_emp   <- latest_snapshot(employment,   emp_rate, "Y15-64") %>%
  rename(employment_pct = emp_rate, emp_date = date)
kpi_act   <- latest_snapshot(activity,     act_rate, "Y15-64") %>%
  rename(activity_pct = act_rate, act_date = date)
kpi_vac   <- latest_snapshot(vacancy_rate, v_rate) %>%
  rename(vacancy_pct = v_rate, vac_date = date)

kpi_table <- tibble(country_name = country_order) %>%
  left_join(kpi_unemp, by = "country_name") %>%
  left_join(kpi_emp,   by = "country_name") %>%
  left_join(kpi_act,   by = "country_name") %>%
  left_join(kpi_vac,   by = "country_name")

# Beveridge view: vacancy rate vs headline unemployment, by quarter
beveridge <- vacancy_rate %>%
  select(country_name, date, v_rate) %>%
  inner_join(
    unemployment %>%
      filter(age == "Y15-74") %>%
      select(country_name, date, u_rate),
    by = c("country_name", "date")
  ) %>%
  mutate(year = year(date))

# NACE share = section employment / sum across sections within country-quarter
# Done per quarter to allow comparison even when absolute totals shift
emp_nace_share <- emp_nace %>%
  group_by(country_name, date) %>%
  mutate(total_q = sum(emp_nace_ths, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(share = if_else(total_q > 0, emp_nace_ths / total_q, NA_real_))

# Foreign-born share of employment = FOR / TOTAL.
# Recent Eurostat vintages of lfsq_egacob have dropped the EU28_FOR /
# NEU28_FOR dimensions, so make pivot_wider robust to columns that may not
# be present in the current data vintage.
cob_share_wide <- emp_cob %>%
  filter(c_birth %in% c("TOTAL", "FOR", "EU28_FOR", "NEU28_FOR")) %>%
  select(country_name, date, c_birth, emp_cob_ths) %>%
  pivot_wider(names_from = c_birth, values_from = emp_cob_ths)

for (col in setdiff(c("TOTAL", "FOR", "EU28_FOR", "NEU28_FOR"),
                    names(cob_share_wide))) {
  cob_share_wide[[col]] <- NA_real_
}

cob_share <- cob_share_wide %>%
  mutate(
    foreign_share = if_else(!is.na(TOTAL) & TOTAL > 0, FOR / TOTAL, NA_real_),
    eu_share      = if_else(!is.na(TOTAL) & TOTAL > 0, EU28_FOR / TOTAL, NA_real_),
    non_eu_share  = if_else(!is.na(TOTAL) & TOTAL > 0, NEU28_FOR / TOTAL, NA_real_)
  )

# Whether the EU vs non-EU foreign-born breakdown is available at all in the
# current cache — controls whether the second panel on the Country of Birth
# tab renders the chart or a brief "not available" note instead.
has_eu_split <- any(!is.na(cob_share$eu_share))

# Shared ggplot pieces
date_scale <- scale_x_date(date_breaks = "3 years", date_labels = "%Y")

bgi_line_theme <- theme_minimal(base_family = "sans") +
  theme(
    plot.title       = element_text(color = "#115780", size = 13),
    plot.caption     = element_text(color = "#6b6b6b", size = 8,
                                    hjust = 0, face = "italic"),
    legend.position  = "bottom",
    legend.title     = element_blank(),
    panel.grid.minor = element_blank()
  )

# Force the country factor order in every plot so legends line up across tabs
order_country <- function(df) {
  df %>% mutate(country_name = factor(country_name, levels = country_order))
}
```

Overview {data-icon="fa-globe"}
=======================================================================

Row {data-height=170}
-----------------------------------------------------------------------

### Latest-quarter snapshot {.value-box-row}

```{r kpi-table}
# Single compact table instead of 28 value boxes — easier to scan across
# countries and indicators in one view.
kpi_table %>%
  transmute(
    Country      = country_name,
    `Unemp date`   = format(unemp_date, "%Y-%m"),
    `Unemployment 15-74 (%)`  = round(unemployment_pct, 1),
    `Employment 15-64 (%)`    = round(employment_pct, 1),
    `Activity 15-64 (%)`      = round(activity_pct, 1),
    `Vacancy rate (%)`        = round(vacancy_pct, 1)
  ) %>%
  datatable(
    rownames = FALSE,
    options  = list(dom = "t", paging = FALSE, ordering = FALSE)
  )
```

Row
-----------------------------------------------------------------------

### Headline unemployment rate (15–74) — all economies

```{r overview-unemployment-chart}
unemployment %>%
  filter(age == "Y15-74") %>%
  order_country() %>%
  ggplot(aes(x = date, y = u_rate, color = country_name)) +
  geom_line(size = 0.7) +
  scale_color_bgi() +
  date_scale +
  labs(
    title   = "Harmonised unemployment rate, 15-74, seasonally adjusted",
    x       = NULL,
    y       = "% of labour force",
    caption = paste(
      "Sources: Eurostat (une_rt_q), ONS, FRED (LRUN74TTUSQ156S).",
      "US and UK reweighted to OECD-harmonised basis;",
      "small methodological differences vs Eurostat remain."
    )
  ) +
  bgi_line_theme
```

Unemployment {data-icon="fa-user-slash"}
=======================================================================

Row
-----------------------------------------------------------------------

### Headline unemployment (15–74)

```{r unemployment-headline}
unemployment %>%
  filter(age == "Y15-74") %>%
  order_country() %>%
  ggplot(aes(x = date, y = u_rate, color = country_name)) +
  geom_line(size = 0.7) +
  scale_color_bgi() +
  date_scale +
  labs(
    title   = "Unemployment rate, 15-74",
    x       = NULL, y = "% of labour force",
    caption = "Sources: Eurostat une_rt_q (EU5), ONS (UK 2021+), FRED LRUN74TTUSQ156S (US). Seasonally adjusted."
  ) +
  bgi_line_theme
```

Row
-----------------------------------------------------------------------

### Change in unemployment rate vs 2019 average (percentage points)

```{r unemployment-vs-2019}
# Originally planned this panel as youth vs prime-age small multiples, but the
# underlying sources don't overlap: Eurostat publishes Y15-74 + Y25-74,
# FRED-OECD (US) publishes Y15-74 + Y_LT25, and ONS (UK) publishes Y15-74
# only — there is no shared age split across all 7 countries. Substituting
# a recovery-from-2019 view that uses the universally-available Y15-74 series.
unemployment_2019_baseline <- unemployment %>%
  filter(age == "Y15-74", year(date) == 2019) %>%
  group_by(country_name) %>%
  summarise(baseline_u = mean(u_rate, na.rm = TRUE), .groups = "drop")

unemployment %>%
  filter(age == "Y15-74", date >= as.Date("2019-01-01")) %>%
  inner_join(unemployment_2019_baseline, by = "country_name") %>%
  mutate(delta_pp = u_rate - baseline_u) %>%
  order_country() %>%
  ggplot(aes(x = date, y = delta_pp, color = country_name)) +
  geom_hline(yintercept = 0, color = "#999999", linewidth = 0.3) +
  geom_line(size = 0.7) +
  scale_color_bgi() +
  date_scale +
  labs(
    title   = "Unemployment rate, change vs 2019 average",
    x       = NULL, y = "Percentage-point change vs 2019",
    caption = paste(
      "Baseline = mean(Y15-74 unemployment, 2019).",
      "Positive = labour market weaker than 2019; negative = tighter than 2019."
    )
  ) +
  bgi_line_theme
```

Employment & Activity {data-icon="fa-briefcase"}
=======================================================================

Row
-----------------------------------------------------------------------

### Employment rate (15–64)

```{r employment-rate}
# Y15-64 used because US (FRED OECD-harmonised) and UK (ONS) only publish
# this age band — Y20-64 would silently drop those two countries.
employment %>%
  filter(age == "Y15-64") %>%
  order_country() %>%
  ggplot(aes(x = date, y = emp_rate, color = country_name)) +
  geom_line(size = 0.7) +
  scale_color_bgi() +
  date_scale +
  labs(
    title   = "Employment rate, 15-64",
    x       = NULL, y = "% of working-age population",
    caption = "Sources: Eurostat lfsi_emp_q, ONS, FRED LREM64TTUSQ156S. Seasonally adjusted."
  ) +
  bgi_line_theme
```

### Activity rate (15–64)

```{r activity-rate}
activity %>%
  filter(age == "Y15-64") %>%
  order_country() %>%
  ggplot(aes(x = date, y = act_rate, color = country_name)) +
  geom_line(size = 0.7) +
  scale_color_bgi() +
  date_scale +
  labs(
    title   = "Labour-force participation, 15-64",
    x       = NULL, y = "% of population",
    caption = "Sources: Eurostat lfsi_emp_q (indic_em=ACT), ONS, FRED LRAC64TTUSQ156S. Seasonally adjusted."
  ) +
  bgi_line_theme
```

Vacancies {data-icon="fa-bullhorn"}
=======================================================================

Row
-----------------------------------------------------------------------

### Vacancy rate (% of total posts, NACE B–S)

```{r vacancy-line}
vacancy_rate %>%
  order_country() %>%
  ggplot(aes(x = date, y = v_rate, color = country_name)) +
  geom_line(size = 0.7) +
  scale_color_bgi() +
  date_scale +
  labs(
    title   = "Job vacancy rate",
    x       = NULL, y = "% of occupied + vacant posts",
    caption = paste(
      "Sources: Eurostat jvs_q_nace2 (EU5 + UK pre-2021), ONS Vacancy Survey (UK 2021+), FRED JTSJOR (US).",
      "Coverage: NACE Rev. 2 sections B-S."
    )
  ) +
  bgi_line_theme
```

Row
-----------------------------------------------------------------------

### Beveridge curve — vacancies vs unemployment, faceted by country

```{r beveridge}
# Each panel traces the historical (u, v) path of one country, with point
# colour graded by year. The orange-green diverging scale gives an intuitive
# old-to-new reading (orange = older, green = newer).
beveridge %>%
  order_country() %>%
  ggplot(aes(x = u_rate, y = v_rate, color = year)) +
  geom_path(alpha = 0.4, linewidth = 0.4) +
  geom_point(size = 1.1) +
  facet_wrap(~ country_name, ncol = 4, scales = "free") +
  scale_color_bgi_og(midpoint = median(beveridge$year, na.rm = TRUE)) +
  labs(
    title   = "Beveridge curve: vacancies vs unemployment by quarter",
    x       = "Unemployment rate (%)",
    y       = "Vacancy rate (%)",
    color   = "Year",
    caption = paste(
      "Sources: vacancy + unemployment series as in adjacent tabs.",
      "Free axes per country — shape and shifts of the curve matter more than absolute levels."
    )
  ) +
  bgi_line_theme +
  theme(legend.position = "right")
```

Industry composition {data-icon="fa-industry"}
=======================================================================

Row
-----------------------------------------------------------------------

### Latest-quarter share of employment by NACE section

```{r emp-nace-bars}
# A grouped bar across countries gets cluttered with 21 sections, so we facet
# by country and let each bar express its country's share within that quarter.
emp_nace_share_latest <- emp_nace_share %>%
  group_by(country_name) %>%
  filter(date == max(date, na.rm = TRUE)) %>%
  ungroup() %>%
  order_country()

emp_nace_share_latest %>%
  mutate(nace_r2 = factor(nace_r2, levels = sort(unique(nace_r2), decreasing = TRUE))) %>%
  ggplot(aes(x = share, y = nace_r2, fill = country_name)) +
  geom_col() +
  facet_wrap(~ country_name, ncol = 4) +
  scale_fill_bgi() +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title   = "Share of employment by NACE Rev. 2 section (latest quarter)",
    x       = "Share of total employment",
    y       = "NACE section",
    caption = paste(
      "Sources: Eurostat lfsq_egan2 (EU5 + UK pre-2021), Nomis NM_130_1 (UK 2021+), BLS CES (US).",
      "Caveat: BLS CES groups O-Q and R-U are split evenly across constituent NACE letters — approximate."
    )
  ) +
  bgi_line_theme +
  theme(legend.position = "none",
        axis.text.y     = element_text(size = 7))
```

Row
-----------------------------------------------------------------------

### Top-4 sectors by latest-quarter share — share over time

```{r emp-nace-top4-trend}
# For each country, identify the four largest NACE sections in the latest
# quarter, then plot their share of employment over time within that country.
top4_per_country <- emp_nace_share %>%
  group_by(country_name) %>%
  filter(date == max(date, na.rm = TRUE)) %>%
  slice_max(share, n = 4, with_ties = FALSE) %>%
  ungroup() %>%
  select(country_name, nace_r2)

emp_nace_share %>%
  inner_join(top4_per_country, by = c("country_name", "nace_r2")) %>%
  order_country() %>%
  ggplot(aes(x = date, y = share, color = nace_r2)) +
  geom_line(size = 0.55) +
  facet_wrap(~ country_name, ncol = 4, scales = "free_y") +
  scale_color_bgi() +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  date_scale +
  labs(
    title   = "Largest 4 NACE sections per country — share over time",
    x       = NULL, y = "Share of total employment",
    color   = "NACE",
    caption = "Top-4 sections identified at the latest available quarter and held fixed across time within each country."
  ) +
  bgi_line_theme
```

Country of birth {data-icon="fa-globe-europe"}
=======================================================================

Row
-----------------------------------------------------------------------

### Foreign-born share of employment over time

```{r cob-foreign-share}
cob_share %>%
  filter(!is.na(foreign_share)) %>%
  order_country() %>%
  ggplot(aes(x = date, y = foreign_share, color = country_name)) +
  geom_line(size = 0.7) +
  scale_color_bgi() +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  date_scale +
  labs(
    title   = "Foreign-born share of employment, ages 15-64",
    x       = NULL, y = "Foreign-born / total employed",
    caption = paste(
      "Sources: Eurostat lfsq_egacob (EU5 + UK pre-2021), CPS via FRED (US).",
      "Pre-2023 Eurostat releases populate Q2 only; intervening quarters appear as gaps.",
      "Caveat: US foreign-born series uses CPS — slightly different concept vs the EU LFS."
    )
  ) +
  bgi_line_theme
```

Row
-----------------------------------------------------------------------

### EU vs non-EU foreign-born share, where available

```{r cob-eu-vs-non-eu, eval=has_eu_split}
# This split exists only in the Eurostat country-of-birth release — US and
# UK post-2021 only carry the aggregate FOR series, so they show empty panels
# here on purpose (preserves the 7-country grid so the eye can spot what's
# missing).
cob_share %>%
  select(country_name, date, eu_share, non_eu_share) %>%
  pivot_longer(c(eu_share, non_eu_share),
               names_to = "origin", values_to = "share") %>%
  filter(!is.na(share)) %>%
  mutate(
    origin = recode(origin,
                    "eu_share"     = "EU-born",
                    "non_eu_share" = "Non-EU-born")
  ) %>%
  order_country() %>%
  ggplot(aes(x = date, y = share, color = origin)) +
  geom_line(size = 0.55) +
  facet_wrap(~ country_name, ncol = 4, scales = "free_y") +
  scale_color_bgi() +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  date_scale +
  labs(
    title   = "Foreign-born share of employment by origin",
    x       = NULL, y = "Share of total employed",
    color   = NULL,
    caption = paste(
      "Source: Eurostat lfsq_egacob — EU28_FOR and NEU28_FOR series.",
      "US and UK (post-2021) only publish the aggregate FOR series and so appear blank in this view."
    )
  ) +
  bgi_line_theme
```

```{r cob-eu-vs-non-eu-missing, eval=!has_eu_split, results='asis'}
# Fallback note when the current cache has no EU28_FOR / NEU28_FOR rows
# (recent Eurostat vintages of lfsq_egacob omit these dimensions).
cat(
  "**EU vs non-EU foreign-born breakdown is not available in the current ",
  "data vintage.** The Eurostat `lfsq_egacob` release backing the rest of ",
  "this tab no longer publishes the `EU28_FOR` / `NEU28_FOR` `c_birth` ",
  "categories — only the aggregate `FOR` (shown above), `NAT`, and ",
  "`TOTAL` series. Refresh `data/eurostat_data_*.RDS` from a vintage that ",
  "still includes the split to re-enable this panel.\n",
  sep = ""
)
```
